---
title: 'GCalignR: An R package for aligning Gas-Chromatography data'
author: Meinolf Ottensmann, Martin A. Stoffel, Barbara Casper, Joseph I. Hoffman
date: "11/1/2015"
output:
  pdf_document:
    fig_caption: yes
    highlight: kate
    keep_tex: yes
    number_sections: yes
  html_document:
    fig_caption: yes
    force_captions: yes
    highlight: pygments
    number_sections: yes
    theme: cerulean
  word_document: default
bibliography: bibliography.bib
---

```{r, echo=FALSE, results='hide'}
  # devtools::install_github("cboettig/knitcitations@v1")
  library(knitcitations)
  cleanbib()
  cite_options(citation_format = "pandoc", check.entries=FALSE)
  library(bibtex)
``` 

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", cache = TRUE) # warning = FALSE
```


# Abstract

Metabolomics approaches such as Gas-Chromatography (GC) are increasingly used by researchers in
ecology and evolution to study the chemical basis of animal communication. However, the statistical
analysis of patterns across individuals requires the alignment of chemical data, which is inherently
noisy due to subtle, random and often unavoidable variation of the GC machine parameters [@pierce2005classification]. We developed an R package, `GCalignR`, which provides easy-to-use
function to align GC data in R and evaluate the quality of the alignment. 

Key-words: GC-MS, chemical communication, olfactory communication, alignment

	
# Introduction

Despite the existence of automated alignment algorithms [e.g. @smith2006xcms; @stein1999integrated; @robinson2007dynamic], most researchers in the relatively young fields of mammalian and avian chemical communication align chromatograms manually
[@charpentier2008smelling; @setchell2010odour; @caspers2011scents; @leclaire2012semiochemical; @theis2013symbiotic] or identify all compounds prior to analysis [@whittaker2010songbird] rather than using automated alignment software. Depending on the
sample size and the number of compounds present in the samples, manual alignment can take up to several weeks and might be biased due to subjective expectations of the researcher. 


## Availability

The latest version of `GCalignR` can be downloaded from GitHub.

```{r, eval = FALSE}
install.packages("devtools")
devtools::install_github("mastoffel/GCalignR")
```

We welcome any contributions or feedback on the package.

## Data accessibility


## References




