---
title: 'Supplementary for "GCalignR: An R package for aligning Gas-Chromatography data"'
author: "Meinolf Ottensmann, Martin A. Stoffel, Barbara Caspers, Joseph I. Hoffman"
date: '`r Sys.Date()`'
output:
  pdf_document:
    fig_caption: yes
    highlight: kate
    keep_tex: yes
    number_sections: yes
  html_document:
    fig_caption: yes
    force_captions: yes
    highlight: pygments
    number_sections: yes
    theme: cerulean
  word_document: default
bibliography: bibliography.bib
---

```{r, echo = FALSE, results = 'hide'}
  # devtools::install_github("cboettig/knitcitations@v1")
  library(knitcitations)
  cleanbib()
  cite_options(citation_format = "pandoc", check.entries = FALSE)
  library(bibtex)
``` 

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", cache = TRUE) # warning = FALSE
```
## Literature review
There are more than 53 000 hits when searching for *gas-chromatography* in Web of Science (24.11.2016). Most of them are issued in journals focussing on analytical sciences such as diverse fields in chemistry, food science or pharmacology. The focus of GCalignR is on applications were not the individual substance is of primary interest but rather patterns or chemical signatures. Hence, a literature review is conducted by considering articles published in the research area of ecology in the past ten years. The aim is to investigate the prevelance of manual alignments of gas-chromatography data. 

### Journals

* Animal Behaviour
* Auk
* Behavioral Ecology
* Behavioral Ecology and Sociobiology
* Ecology
* Plos ONE
* Zoology

## References